---
title: "Genome Informatics"
author: "Loryn Young"
date: "2/19/2022"
output: pdf_document
---

# Section 1. Proportion of G/G in a population 

Downloaded a CSV file from Ensemble < https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39869101-39921102;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel > 

Here we read this CSV file 


```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Lets now dig into this further. 

## Section 4: Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

How many samples do we have? 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```


```{r}
table(expr$geno)
```


```{r}
library(ggplot2)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

Lets make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```
Based on the boxplot, we can infer that there is a higher expression value of A/A compared to G/G. Having a G/G in this location is associated with having reduced expression of this gene. We know that the SNP effects the expression of ORMDL3 because the genotypes result in different expression levels, and different SNPs are associated with different genotypes. 
