---
title: "Mini Project"
author: "Loryn Young"
date: "2/8/2022"
output: pdf_document
---

1. Exploratory data analysis

```{r}
fna.data <- "https://bioboot.github.io/bimm143_S20/class-material/WisconsinCancer.csv"
```

```{r}
wisc.df <- read.csv(fna.data, row.names=1)
wisc.df
```

```{r}
wisc.data <- wisc.df[,-1]
wisc.data
```

```{r}
diagnosis <- wisc.df$diagnosis
diagnosis <- as.factor(wisc.df[,1])
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
```

212 observations have a malignant diagnosis

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```

2. Principal Component Analysis 

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, scale.=TRUE)
summary(wisc.pr)
```


> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

(0.4427) + (0.1897) + (0.09393) = 0.72363

3 principal components are required to describe at least 70% of the original variance in the data. 


> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

(0.4427) + (0.1897) + (0.09393) + (0.06602) + (0.05496) + (0.04025) + (0.02251) = 0.91

7 principal components are required to describe at least 90% of the original variance in the data. 


#Interpreting PCA results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is super difficult to understand. It looks like all of the red data is coming from one center point of the black data. We will need to make another plot to make sense of this result. 

```{r}
plot(wisc.pr$x[,1:2], col=(diagnosis), xlab="PC1", ylab="PC2")
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

There plot shows a clear distinction between malignant and benign tumors, specifically along PC1

```{r}
plot(wisc.pr$x[,c(1,3)], col=(diagnosis), xlab="PC1", ylab="PC3")
```
```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis
```

```{r}
library(ggplot2)
ggplot(df) + aes(PC1, PC2, col=diagnosis) + geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var / sum(pr.var)
```

```{r}
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
#install.packages("factoextra")

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation[,(colnames(wisc.data) == "concave.points_mean")]
```


> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
pc.min <- 0
index <- 1
while (pc.min < 0.8) {
  pc.min <- pc.min + pve[index]
  index = index + 1
}
paste(index-1, "principal components are required to explain", round(pc.min*100, 1), "% of the data.")
```

3. Hierarchical Clustering 

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method="complete")
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

Height = 19

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=10)
table(wisc.hclust.clusters, diagnosis)
```

When the tree is cut into 10 clusters, it is very easy to recognize which clusters correspond to either malignant or benign cells in the cluster vs. diagnosis match. 

#cut back into 4 clusters
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
```

```{r}
data.dist <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(data.dist, method="ward.D2")
plot(wisc.pr.hclust)
```
> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

"ward.D2" gives me my favorite results for the same data.dist dataset because it is the easiest method for me to pull information from. This is because there is a distinct top and bottom of the tree that points gather at, whereas with some other methods it is harder to understand where data points are coming from.

4. OPTIONAL: K-means clustering

```{r}
wisc.km <- kmeans(scale(wisc.data), centers=2, nstart=20)
table(wisc.km$cluster, diagnosis)
```

> Q14. How well does k-means separate the two diagnoses? How does it compare to your hclust results?

K-means does not separate the two diagnoses as well as the hclust results, but the results are still very similar! Overall the two diagnoses are still separated pretty well. 

5. Combining methods

```{r}
pc.min <- 0
index <- 1
while (pc.min < 0.9) {
  pc.min <- pc.min + pve[index]
  index = index + 1
}
paste(index-1, "principal components are required to explain", round(pc.min*100, 1), "% of the data.")
```


```{r}
data.dist <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(data.dist, method="ward.D2")
plot(wisc.pr.hclust)
```
```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
wisc.pr.hclust <- hclust(data.dist, method="ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
``` 

> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

The newly created model separates the two diagnosis well, it is easy to see which cluster corresponds to which diagnosis. 

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

The k-means and hierarchial clustering models I created do a fair job of separating the diagnoses, because I can clearly distinguish between the corresponding cells for the cluster. The data is very similar to the output of the vector containing the actual diagnoses. 

```{r}
table(wisc.km$cluster, diagnosis)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Best Specificity: k-means
k-means: (343/357) = 0.96
wisc.pr.hclust.clusters: (329/357) = 0.92

Best Sensitivity: hclust
k-means: (175/212) = 0.82
wisc.pr.hclust.clusters: (188/212) = 0.89


7. Prediction 

#url <- "new_samples.csv"
```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

Based on our results, we should prioritize patient 1 because they are farther right on the x-axis (PC1) and the points are clustered very closely. 
