---
title: 'Week 5: R functions'
author: "Loryn Young"
date: "2/6/2022"
output: pdf_document
---
```{r}
plot(1:10)
```
> Q1. . Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

```
```{r}
mean(student1[-which.min(student1)])

```

```{r}
x <- student2
is.na(x)
which( is.na(x))
```

```{r}
x[is.na(x)] <- 0
x
mean(x)
```

```{r}
x[is.na(x)] <- 0
mean( x[-which.min(x)])
```
```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean( x[-which.min(x)])
```
```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean( x[-which.min(x)])
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)
```
```{r}
#' Calculate the average score for a vector of student scores dropping the lowest score.
#' Missing values will be treated as zero. 
#'
#' @param x A numeric vector of homework scores
#'
#' @return Average score
#' @export
#'
#' @examples
#' student <- c(100, NA, 90, 97)
#' grade(student)
#' 
grade <- function(x) {
  #mask NA with zero
  #Treat missing values as zero
  x[is.na(x)] <- 0
  #Exclude lowest score from mean
  mean( x[-which.min(x)])
}
```
```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names=1)
```

```{r}
apply(gradebook, 1, grade)
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts]

```{r}
results <- apply(gradebook, 1, grade)
sort(results, decreasing = TRUE)
```

```{r}
which.max(results)
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
gradebook
```
```{r}
med.scores <- apply(gradebook, 2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```
```{r}
boxplot(gradebook)
```
> Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
masked.gradebook <- gradebook
masked.gradebook [ is.na(masked.gradebook)] <- 0
masked.gradebook
```
```{r}
cor(results, masked.gradebook$hw5)
```
```{r}
apply(masked.gradebook, 2, cor, x=results)
```


> Q5. Make sure you save your Rmarkdown document and can click the “Knit” button to generate
a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]
